---
title: "How common are trout in Driftless streams"
author: "Bryan Maitland"
date: "29 July 2021"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, 
                      dev = "ragg_png", res = 500, retina = 1)
```

## Overview

What is the percentage of Driftless area streams in which Brook, Brown, or Tiger Trout are present?

```{r prep, message=FALSE, warning=FALSE}
library(tidyverse)
library(here)
library(wdnr.fmdb)
set_fmdb_credentials()
library(rgdal)  
library(sf)

theme_set(theme_void(base_family = "sans"))
```


## 1. Data

### FMDB surveys and efforts

Download all surveys from 1994-2020 on streams using backpack and stream shockers, retaining only surveys that are complete and proofed:

```{r pull-save-raw-survey-effort-data}
# # params for pulling surveys and efforts
# yrs <- list(1994:1999,2000:2005,2006:2011,2012:2016,2017:2020)
# waterbody_types <- c( "wadable_stream", "non-wadable_stream", "stream")
# gear_types <- c("stream_shocker","backpack_shocker")
# 
# # pull surveys
# df_surveys_raw <- 
#   yrs %>% 
#   map_df(~get_fmdb_surveys(year = ., waterbody_type = waterbody_types))
# 
# # pulls efforts
# df_efforts_raw <- 
#   yrs %>% 
#   map_df(~get_fmdb_efforts(year = ., waterbody_type = waterbody_types, gear = gear_types))
# 
# # clean up
# rm(yrs); rm(gear_types); rm(waterbody_types)
# 
# # wrtie to file
# write_rds(df_surveys_raw, here("data", "surveys_raw_20210803.rds"))
# write_rds(df_efforts_raw, here("data", "efforts_raw_20210803.rds"))

# load
df_surveys_raw <- read_rds(here("data", "surveys_raw_20210803.rds"))
df_efforts_raw <- read_rds(here("data", "efforts_raw_20210803.rds"))
```

```{r filter-for-proofed-surveys-w-trout-targets}
# # check surveys with no match in the efforts (which are for electofishing only)
# df_surveys_raw %>% 
#   anti_join(df_efforts_raw, by = "survey.seq.no") 
# # these are IBI surveys or those with efforts that are "multiple_gear_types"

# filter for unique surveys in the efforts data
df_surveys <- semi_join(df_surveys_raw, df_efforts_raw, by = "survey.seq.no")

# filter for proofed survey data
targ.survs <- c(
  "data_entry_complete_and_proofed",
  "historical_data_complete_and_proofed",
  "historical_data_entry_complete",
  "historical_data_load_status_unknown")
df_surveys <- filter(df_surveys, survey.status %in% targ.survs)


# filter efforts
targ.spp <- c(
  "all_species","gamefish_species","gamefish_panfish",
  "trout_spp","brown_trout","brook_trout","rainbow_trout")

df_efforts <- 
  df_efforts_raw %>% 
  semi_join(df_surveys, by = "survey.seq.no") %>%
  filter(target.species %in% targ.spp | secondary.target.species %in% targ.spp) %>% 
  filter(site.seq.no != 315) 

# surveys again by the cleaned efforts
df_surveys <- semi_join(df_surveys, df_efforts, by = "survey.seq.no")

# clean up 
rm(df_surveys_raw);rm(df_efforts_raw)
rm(targ.survs); rm(targ.spp)
```


### clip data to driftless region:

```{r load-driftless-polys, cache=TRUE, results = 'hide', echo=FALSE}
# load driftless polygon
driftless <- 
  here("data", "ecoregions","wi_eco_l3.shp") %>% 
  st_read() %>% 
  st_transform(crs = 3071) %>% 
  filter(US_L3CODE == "52")

# load classified trout stream lines
lines <- 
  here("data", "classified_trout_streams","Classified_Trout_Stream_Lines.shp") %>% 
  st_read() %>% 
  st_transform(crs = 3071) 
```

```{r clip-surveys-effs-to-driftless, cache=TRUE}
# clip the classified trout streams to driftless region
lines_drift <- 
  lines %>% 
  st_intersection(driftless)

# make surveys spatial and clip to driftless area
df_surveys_drift <- 
  df_surveys %>% 
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>% 
  st_transform(crs = 3071) %>% 
  st_intersection(driftless)

# make a simple xwalk for wbic and trout class
xwalk_wbic_troutclass <- 
  lines_drift %>% 
  st_drop_geometry() %>% 
  as_tibble() %>% 
  distinct(WBIC, .keep_all = TRUE) %>% 
  select(WBIC, TROUT_CLAS)

# filter efforts for driftless surveys
df_efforts_drift <- semi_join(df_efforts, df_surveys_drift, by = "survey.seq.no")
  
# clean up 
rm(df_surveys); rm(df_efforts)
```


### join surevys sites to stream lines

```{r join-sites-to-lines, cache=TRUE}

# join trout class information to each survey
df_surveys_drift_class <- 
  df_surveys_drift %>% 
  left_join(xwalk_wbic_troutclass, by = c("wbic" = "WBIC")) %>% 
  as_tibble() %>% 
  mutate(TROUT_CLAS = if_else(is.na(TROUT_CLAS), "Unclassified", TROUT_CLAS)) 

# count of surveys on different class streams
df_surveys_drift_class %>% 
  count(TROUT_CLAS)

# isolate unique survey site locations in driftless
# df_surveys_site_locs <- 
#   df_surveys_drift_class %>% 
#   distinct(site.seq.no, .keep_all = TRUE)
```

### download fishraw data

```{r pull-raw-fish, echo=FALSE}
# set chunks of visits seqs for pulling fish data
vs <- unique(df_efforts_drift$visit.fish.seq.no)
vs_chunks <- split(vs, ceiling(seq_along(vs)/1000))

df_fish_drift <-
  vs_chunks %>%
  map_df(~get_fmdb_fishraw(visit_seq = .))

# clean up
rm(vs); rm(vs_chunks);
# rm(length_deleted_rows); rm(length_warning_rows)

# write to file
write_rds(df_fish_drift, here("data", "fish_raw_20210804.rds"))

# load
df_fish_drift <- read_rds(here("data", "fish_raw_20210804.rds"))

```

```{r df-of-zeros, echo=FALSE}
df_trout_0s <-
  bind_rows(
    # all species and game fish targets
    df_fish_drift %>%
      filter(target.species %in% c("all_species","gamefish_species"),
             species == "no_fish_captured") %>%
      mutate(
        species = "brook_trout",
        number.of.fish = 0,
        ),
    df_fish_drift %>%
      filter(target.species %in% c("all_species","gamefish_species"),
             species == "no_fish_captured") %>%
      mutate(
        species = "brown_trout",
        number.of.fish = 0,
        ),
    # brook trout targets
    df_fish_drift %>%
      filter(target.species %in% c("brook_trout"),
             species == "no_fish_captured") %>%
      mutate(
        species = "brook_trout",
        number.of.fish = 0,
        ),
    # brown trout targets
    df_fish_drift %>%
      filter(target.species %in% c("brown_trout"),
             species == "no_fish_captured") %>%
      mutate(
        species = "brown_trout",
        number.of.fish = 0,
        )
  )
```

```{r add-zeros-and-subset-trout}
df_fish_drift_trout <- 
  df_fish_drift %>% 
  filter(!species == "no_fish_captured") %>% bind_rows(df_trout_0s) %>%
  mutate(species = if_else(species == "tiger_trout_(i21_x_i22)", "tiger_trout", species)) %>% 
  filter(species %in% c("brook_trout","brown_trout","tiger_trout")) %>% 
  filter(!str_detect(wbic, "^999"))

rm(df_trout_0s)
```


```{r missing}

# 3 surveys do not have any fish data
# the other 27 missing wbics get deleted when fioltering for trout
missing_fish_wbics <- 
  anti_join(
    df_surveys_drift_class %>% 
      filter(TROUT_CLAS %in% c("CLASS I", "CLASS II", "CLASS III")) %>% 
      distinct(wbic), 
    df_fish_drift_trout %>% 
      left_join(xwalk_wbic_troutclass , by = c("wbic"="WBIC")) %>% 
      filter(TROUT_CLAS %in% c("CLASS I", "CLASS II", "CLASS III")) %>% 
      distinct(wbic), 
    by = "wbic") %>% 
  pull()


missing_fish_survs <-
  df_surveys_drift_class %>% 
  filter(wbic %in% missing_fish_wbics)

missing_efforts <- 
  df_efforts_drift %>% 
  filter(wbic %in% missing_fish_wbics)


missing_efforts$visit.fish.seq.no %in% vs

missing_data <- 
  get_fmdb_fishraw(visit_seq = missing_efforts$visit.fish.seq.no)
  
missing_data %>% 
  filter(wbic %in% missing_fish_wbics)

```


## 2. Count of surveys and streams (wbics)

### count of surveyed wbics in which trout are present:

```{r convert-to-pa}

# df_fish_drift_pa <-
#   df_fish_drift %>% 
#   filter(!str_detect(wbic, "^999")) %>% 
#   group_by(wbic, survey.seq.no, species) %>% 
#   summarise(n = sum((number.of.fish)), .groups = "drop") %>% 
#   mutate(n = if_else(n > 1, 1, 0)) %>% 
#   pivot_wider(names_from = species, values_from = n, values_fill = 0) %>%  
#   mutate(
#     no_trout = if_else(brook_trout==0 & brown_trout==0, 1, 0), 
#     symp = if_else(brook_trout==1 & brown_trout==1, 1, 0), 
#     bkt_only = if_else(brook_trout==1 & brown_trout==0, 1, 0), 
#     bnt_only = if_else(brook_trout==0 & brown_trout==1, 1, 0)
#     ) %>% 
#   pivot_longer(cols = 3:9, names_to = "species", values_to = "present") %>% 
#   left_join(df_surveys_drift_class %>%
#               select(survey.seq.no, TROUT_CLAS, geometry), 
#             by = "survey.seq.no") %>% 
#   st_as_sf()
# df_fish_drift_pa
# 
# df_fish_drift_pa %>% 
#   group_by(wbic, species) %>% 
#   summarise(n = sum(present), .groups = "drop") %>% 
#   mutate(n = if_else(n > 1, 1, 0))

df_fish_drift_trout_pa <-
  df_fish_drift_trout %>% 
  group_by(wbic, species) %>% 
  summarise(n = sum((number.of.fish)), .groups = "drop") %>% 
  mutate(n = if_else(n > 1, 1, 0)) %>% 
  pivot_wider(names_from = species, values_from = n, values_fill = 0) %>%  
  mutate(
    no_trout = if_else(brook_trout==0 & brown_trout==0, 1, 0),
    symp = if_else(brook_trout==1 & brown_trout==1, 1, 0),
    bkt_only = if_else(brook_trout==1 & brown_trout==0, 1, 0),
    bnt_only = if_else(brook_trout==0 & brown_trout==1, 1, 0)
    ) %>%
  pivot_longer(cols = 2:8, names_to = "species", values_to = "present") %>% 
  left_join(df_surveys_drift_class %>%
              select(wbic, TROUT_CLAS), 
            by = "wbic")

```

```{r count-wbics-with-trout-present}
n.wbics.bnt <- 
  df_fish_drift_trout_pa %>% 
  filter(species == "brown_trout" & present == 1) %>% 
  pull(wbic) %>% unique() %>% length()

n.wbics.bnt.123 <- 
  df_fish_drift_trout_pa %>% 
  filter(TROUT_CLAS %in% c("CLASS I","CLASS II","CLASS III")) %>% 
  filter(species == "brown_trout" & present == 1) %>% 
  pull(wbic) %>% unique() %>% length()

n.wbics.bnt.12 <- 
  df_fish_drift_trout_pa %>% 
  filter(TROUT_CLAS %in% c("CLASS I","CLASS II")) %>% 
  filter(species == "brown_trout" & present == 1) %>% 
  pull(wbic) %>% unique() %>% length()

n.wbics.bkt <- 
  df_fish_drift_trout_pa %>% 
  filter(species == "brook_trout" & present == 1) %>% 
  pull(wbic) %>% unique() %>% length()

n.wbics.bkt.123 <- 
  df_fish_drift_trout_pa %>% 
  filter(TROUT_CLAS %in% c("CLASS I","CLASS II","CLASS III")) %>% 
  filter(species == "brook_trout" & present == 1) %>% 
  pull(wbic) %>% unique() %>% length()

n.wbics.bkt.12 <- 
  df_fish_drift_trout_pa %>% 
  filter(TROUT_CLAS %in% c("CLASS I","CLASS II")) %>% 
  filter(species == "brook_trout" & present == 1) %>% 
  pull(wbic) %>% unique() %>% length()


n.wbics.tgt <- 
  df_fish_drift_trout_pa %>% 
  filter(species == "tiger_trout" & present == 1) %>% 
  pull(wbic) %>% unique() %>% length()

n.wbics.tgt.123 <- 
  df_fish_drift_trout_pa %>% 
  filter(TROUT_CLAS %in% c("CLASS I","CLASS II","CLASS III")) %>% 
  filter(species == "tiger_trout" & present == 1) %>% 
  pull(wbic) %>% unique() %>% length()

n.wbics.tgt.12 <- 
  df_fish_drift_trout_pa %>% 
  filter(TROUT_CLAS %in% c("CLASS I","CLASS II")) %>% 
  filter(species == "tiger_trout" & present == 1) %>% 
  pull(wbic) %>% unique() %>% length()


n.wbics.smp <- 
  df_fish_drift_trout_pa %>% 
  filter(species == "symp" & present == 1) %>% 
  pull(wbic) %>% unique() %>% length()

n.wbics.smp.123 <- 
  df_fish_drift_trout_pa %>% 
  filter(TROUT_CLAS %in% c("CLASS I","CLASS II","CLASS III")) %>% 
  filter(species == "symp" & present == 1) %>% 
  pull(wbic) %>% unique() %>% length()

n.wbics.smp.12 <- 
  df_fish_drift_trout_pa %>% 
  filter(TROUT_CLAS %in% c("CLASS I","CLASS II")) %>% 
  filter(species == "symp" & present == 1) %>% 
  pull(wbic) %>% unique() %>% length()


n.wbics.smp.123 <- 
  df_fish_drift_trout_pa %>% 
  filter(TROUT_CLAS %in% c("CLASS I","CLASS II","CLASS III")) %>% 
  filter(species == "symp" & present == 1) %>% 
  pull(wbic) %>% unique() %>% length()


n.wbics.bkt.only.123 <- 
  df_fish_drift_trout_pa %>% 
  filter(TROUT_CLAS %in% c("CLASS I","CLASS II","CLASS III")) %>% 
  filter(species == "bkt_only" & present == 1) %>% 
  pull(wbic) %>% unique() %>% length()

n.wbics.bnt.only.123 <- 
  df_fish_drift_trout_pa %>% 
  filter(TROUT_CLAS %in% c("CLASS I","CLASS II","CLASS III")) %>% 
  filter(species == "bnt_only" & present == 1) %>% 
  pull(wbic) %>% unique() %>% length()

n.wbics.notrout.123 <- 
  df_fish_drift_trout_pa %>% 
  filter(TROUT_CLAS %in% c("CLASS I","CLASS II","CLASS III")) %>% 
  filter(species == "no_trout" & present == 1) %>% 
  pull(wbic) %>% unique() %>% length()

```

#### number of surveyed wbics

```{r count-surveyed-wbics}
# # unique surveys and wbics in driftles surveys data
# n.surveyed.wbics <- length(unique(df_surveys_drift_class$wbic))
# 
# # unique surveys and wbics in driftles surveys data class 1-3
# n.surveyed.wbics.class123 <- 
#   df_surveys_drift_class %>% 
#   filter(TROUT_CLAS %in% c("CLASS I", "CLASS II", "CLASS III")) %>% 
#   summarise(across(c(wbic), ~ length(unique(.x)))) %>% 
#   pull()
# 
# # unique surveys and wbics in driftles surveys data class 1-2
# n.surveyed.wbics.class12 <- 
#   df_surveys_drift_class %>% 
#   filter(TROUT_CLAS %in% c("CLASS I", "CLASS II")) %>% 
#   summarise(across(c(wbic), ~ length(unique(.x)))) %>% 
#   pull()


# unique surveys and wbics in driftles surveys data class 1-3
n.fish.wbics <- 
  df_fish_drift_trout_pa %>% 
  summarise(across(c(wbic), ~ length(unique(.x)))) %>% 
  pull()

# unique surveys and wbics in driftles surveys data class 1-3
n.fish.wbics.class123 <- 
  df_fish_drift_trout_pa %>% 
  filter(TROUT_CLAS %in% c("CLASS I", "CLASS II", "CLASS III")) %>% 
  summarise(across(c(wbic), ~ length(unique(.x)))) %>% 
  pull()

# unique surveys and wbics in driftles surveys data class 1-2
n.fish.wbics.class12 <- 
  df_fish_drift_trout_pa %>% 
  filter(TROUT_CLAS %in% c("CLASS I", "CLASS II")) %>% 
  summarise(across(c(wbic), ~ length(unique(.x)))) %>% 
  pull()

```



## 4. calculate percentages

```{r}
# (percent.wbics.bnt <- round((n.wbics.bnt/n.surveyed.wbics) * 100, digits = 2))
(percent.wbics.bnt.123 <- round((n.wbics.bnt.123/n.fish.wbics.class123) * 100, digits = 2))
# (percent.wbics.bnt.12 <- round((n.wbics.bnt.12/n.fish.wbics.class12) * 100, digits = 2))

# (percent.wbics.bkt <- round((n.wbics.bkt/n.surveyed.wbics) * 100, digits = 2))
(percent.wbics.bkt.123 <- round((n.wbics.bkt.123/n.fish.wbics.class123) * 100, digits = 2))
# (percent.wbics.bkt.12 <- round((n.wbics.bkt.12/n.fish.wbics.class12) * 100, digits = 2))

# (percent.wbics.tgt <- round((n.wbics.tgt/n.surveyed.wbics) * 100, digits = 2))
(percent.wbics.tgt.123 <- round((n.wbics.tgt.123/n.fish.wbics.class123) * 100, digits = 2))
# (percent.wbics.tgt.12 <- round((n.wbics.tgt.12/n.fish.wbics.class12) * 100, digits = 2))

# (percent.wbics.smp <- round((n.wbics.smp/n.surveyed.wbics) * 100, digits = 2))
(percent.wbics.smp.123 <- round((n.wbics.smp.123/n.fish.wbics.class123) * 100, digits = 2))
# (percent.wbics.smp.12 <- round((n.wbics.smp.12/n.surveyed.wbics.class12) * 100, digits = 2))

(percent.wbics.bkt.only <- round((n.wbics.bkt.only.123/n.fish.wbics.class123) * 100, digits = 2))

(percent.wbics.bnt.only <- round((n.wbics.bnt.only.123/n.fish.wbics.class123) * 100, digits = 2))

(percent.wbics.notrout <- round((n.wbics.notrout.123/n.fish.wbics.class123) * 100, digits = 2))


percent.wbics.bkt.only + percent.wbics.bnt.only + percent.wbics.smp.123 + percent.wbics.notrout

```

```{r}

```


## Plot

```{r plot, fig.width=12, fig.height=5}
panel_labels <- c(
  `bkt_only` = glue::glue("BKT Only ({percent.wbics.bkt.only}%)"),
  `bnt_only` = glue::glue("BNT ONLY ({percent.wbics.bnt.only}%)"),
  `no_trout` = glue::glue("No Trout ({percent.wbics.notrout}%)"),

  `brook_trout` = glue::glue("Brook Trout ({percent.wbics.bkt.123}%)"),
  `brown_trout` = glue::glue("Brown Trout ({percent.wbics.bnt.123}%)"),
  `symp` = glue::glue("BKT/BNT Sympatry ({percent.wbics.smp.123}%)"),
  `tiger_trout` = glue::glue("Tiger Trout ({percent.wbics.tgt.123}%)")
  )

ggplot() + 
  geom_sf(data = driftless, fill = "white") +
  geom_sf(data = lines_drift, aes(color = TROUT_CLAS), show.legend = "line") +
  geom_sf(data = df_fish_drift_pa %>% filter(present == 1), aes(fill = TROUT_CLAS), 
          show.legend = "point", size = 1, alpha = 0.5, shape = 21) + 
  scale_colour_manual(
    values = c("#53DC4D", "#00C5ff", "#004DA8"), 
    guide = guide_legend(
      override.aes = list(linetype = c("solid", "solid","solid"), 
                          shape = c(NA, NA, NA),
                          size = c(1,1,1)))
    ) +
  scale_fill_manual(
    values = c("#53DC4D", "#00C5ff", "#004DA8", "grey"), 
    guide = guide_legend(
      override.aes = list(linetype = "blank", shape = 21, size = 2, alpha = 1))
    ) +
  facet_wrap(vars(species), labeller = as_labeller(panel_labels), nrow = 1) +
  labs(
    title = "How common are trout in Driftless Area streams?",
    # title ="Percent of all classified and surveyed streams with trout present:",
    subtitle ="Percent of surveyed trout streams with trout present:",
    caption = "Visualization: Bryan Maitland • Data: WI DNR Fisheries Managment (https://dnr.wisconsin.gov/)", 
    color = "Stream Class", 
    fill = "Trout Presence") + 
  theme(
    plot.title = element_text(hjust=0.5, size=18, face = "bold", margin=margin(0,0,5,0)), 
    plot.subtitle = element_text(hjust=0.5, size=12, margin=margin(0,0,15,0)), 
    plot.caption = element_text(hjust = .5,size = 10, margin = margin(35, 0, 0, 0)), 
    strip.text = element_text(face = "bold")
    )
```

```{r save-and-convert}
path <- here::here("plots", "driftless_trout_presence")

ggsave(glue::glue("{path}.pdf"), width = 15, height = 5, device = cairo_pdf)

pdftools::pdf_convert(pdf = glue::glue("{path}.pdf"),
            filenames = glue::glue("{path}.png"),
            format = "png", dpi = 300)
```

```{r session}
Sys.time()
git2r::repository()
sessionInfo()
```